---
title: "STATS 412 - Final Project - Data Cleaning"
author: "Betty Hwang, Mariano Aloiso"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
filename <- "data/HCMST_2017_to_2022_small_public_version.rdata"
HCMSTData <- get(load(filename))
dim(HCMSTData)
```

Delete all the columns in HCMSTData from the 2022 survey
```{r}
HCMSTData <- HCMSTData[, !grepl("^w3", names(HCMSTData))]
```

Delete all the columns that start with IPSOS KnowledgePanel surveys
```{r}
HCMSTData <- HCMSTData[, !grepl("^p", names(HCMSTData))]
```

Keep only the following columns from the 2020 survey. Delete
everything else that starts with w2
```{r}
relationStatusColumns <- c("w2_section", "W2_rel_qual_reduced")
w2Columns <- names(HCMSTData)[grepl("^w2", names(HCMSTData)) &
                                !names(HCMSTData) %in% relationStatusColumns]
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% w2Columns]
```

Remove people that are not in the 2020 survey
```{r}
HCMSTData <- HCMSTData[!is.na(HCMSTData$w2_section),]
```

w2_section will be the target variable. it is a categorical variable that
indicates whether the subject is still in a relationship in wave 2. We 
transform it into a binary variable (1: still in a relationship, 0: not in a relationship).
"A" (still married) and "B" (still partnered) are considered still in a relationship.
```{r}
HCMSTData$w2_section <- ifelse(HCMSTData$w2_section %in% c(1, 3), 1, 0)
```

# How They Met
Question 24, 25 and 26 asks subjects how they met. There are around 40 possible responses as
this was a free text field. We will group the responses into 6 categories and analyze
this variable independently.
1. Friends
2. Family
3. Online
4. Social Events
5. Work
6. School

Group responses into categories
```{r}
q24_friends <- c(
  "w1_q24_R_friend", "w1_q24_P_friend", "w1_q24_I_friend",
  "w1_q24_met_through_friend", "w1_q24_met_through_as_nghbrs",
  "w1_q24_I_neighbor", "w1_q24_P_neighbor", "w1_q24_R_neighbor",
  "w1_q24_P_sig_other", "w1_q24_R_sig_other", "w1_q24_I_sig_other"
)

q24_family <- c(
  "w1_q24_R_family", "w1_q24_P_family", "w1_q24_I_family",
  "w1_q24_met_through_family"
)

q24_online <- c(
  "w1_q24_internet_other", "w1_q24_internet_dating",
  "w1_q24_internet_soc_network", "w1_q24_internet_game",
  "w1_q24_internet_chat", "w1_q24_internet_org",
  "w1_q24_met_online", "w1_q24_metonline_no_phone_apps",
  "w1_how_met_online", "w1_q32_met_online_phone_apps"
)

q24_social_events <- c(
  "w1_q24_church", "w1_q24_vol_org", "w1_q24_customer",
  "w1_q24_bar_restaurant", "w1_q24_party", "w1_q24_blind_date",
  "w1_q24_vacation", "w1_q24_singles_serve_nonint",
  "w1_q24_public"
)

q24_work <- c(
  "w1_q24_R_cowork", "w1_q24_P_cowork", "w1_q24_I_cowork",
  "w1_q24_work_neighbors", "w1_q24_met_as_through_cowork",
  "w1_q24_business_trip", "w1_q24_mil"
)

q24_school <- c("w1_q24_school", "w1_q24_college", "w1_q25", "w1_q26")

howTheyMetColumns <- c(q24_friends, q24_family, q24_online, q24_social_events,
                       q24_work, q24_school)
```

Subset only the q24 columns along with the w2_section column
```{r}
howTheyMet <- HCMSTData[, c(howTheyMetColumns, "w2_section", "w1_partnership_status")]

for (i in 1:length(q24_friends)) {
  howTheyMet$friends <- ifelse(rowSums(howTheyMet[, q24_friends]) > 0, 1, 0)
}

for (i in 1:length(q24_family)) {
  howTheyMet$family <- ifelse(rowSums(howTheyMet[, q24_family]) > 0, 1, 0)
}

for (i in 1:length(q24_online)) {
  howTheyMet$online <- ifelse(rowSums(howTheyMet[, q24_online]) > 0, 1, 0)
}

for (i in 1:length(q24_social_events)) {
  howTheyMet$social_events <- ifelse(rowSums(howTheyMet[, q24_social_events]) > 0, 1, 0)
}

for (i in 1:length(q24_work)) {
  howTheyMet$work <- ifelse(rowSums(howTheyMet[, q24_work]) > 0, 1, 0)
}

for (i in 1:length(q24_school)) {
  howTheyMet$school <- ifelse(rowSums(howTheyMet[, q24_school]) > 0, 1, 0)
}

howTheyMet <- howTheyMet[, !(names(howTheyMet) %in% c(howTheyMetColumns))]
```

Save the answers to question 24 data in a different dataset.
```{r}
howTheyMet[howTheyMet == -1] <- 0
howTheyMet[is.na(howTheyMet)] <- 0
write.csv(howTheyMet, file = "data/howTheyMet.csv", row.names = FALSE)
```

Drop the q24 columns from the HCMSTData dataframe.
Also drop q25-28, q32 and w1_friend_connect columns as they are related to question 24.
```{r}
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% howTheyMetColumns]
HCMSTData <- HCMSTData[, !grepl("w1_q24", names(HCMSTData))]
questions25to28 <- c("w1_q25", "w1_q26", "w1_q27", "w1_q28", "w1_q32")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% questions25to28]
friendConnectColumns <- c("w1_friend_connect_1_all", "w1_friend_connect_2_all",
                          "w1_friend_connect_3_all", "w1_friend_connect_4_all")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% friendConnectColumns]
```


# Data Collection
Remove features about how the data was collected
```{r}
surveyColumns <- c("w1_speed_flag", "w1_took_the_survey", "w1_weight_combo",
                   "w1_consent", "w1_duration")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% surveyColumns]
```

# Correlation Analysis

Due to the number of features, we do not want to overlook feature pairs that
encode the same data. To find these features, find all variable pairs that
have a correlation of 1 or -1. 
```{r}
correlation_matrix <- cor(HCMSTData)

highly_correlated <- which(upper.tri(correlation_matrix, diag = FALSE),
                           arr.ind = TRUE)
perfectly_correlated_pairs <- subset(highly_correlated,
                        abs(correlation_matrix[highly_correlated]) >= 0.99)

# Display highly correlated variable pairs
for (i in 1:nrow(perfectly_correlated_pairs)) {
  var1 <- rownames(correlation_matrix)[perfectly_correlated_pairs[i, 1]]
  var2 <- colnames(correlation_matrix)[perfectly_correlated_pairs[i, 2]]
  correlation <- correlation_matrix[perfectly_correlated_pairs[i, 1], perfectly_correlated_pairs[i, 2]]
  cat(var1, "and", var2, "are perfectly correlated \n")
}
```

We will remove one of the features from each pair.
Remove w1_female and w1_weight_combo_freqwt
```{r}
variablePairs <- c("w1_weight_combo_freqwt", "w1_female")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% variablePairs]
```

This step also revealed that there is a repeated variable for race, encoded
as w1_pprace* and w1_ppRace_*. The features with an underscore before the number
were self-reported by the participant, while the features without an underscore
came from IPSOS KnowledgePanel. We will remove the IPSOS KnowledgePanel features
since it has a wider range of options (15 options vs 6).
```{r}
raceColumnsIPSOS <- grep("^w1_pprace", names(HCMSTData), value = FALSE)
HCMSTData <- HCMSTData[, -raceColumnsIPSOS]
```

Remove columns "w1w2_sex_freq_diff" and "change_in_rel_quality_w1w2" from the data as
they are calculated using future values
```{r}
futureValuesColumns <- c("w1w2_sex_freq_diff", "change_in_rel_quality_w1w2")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% futureValuesColumns]
```

Remove question 21 from the dataframe (columns that contain Q21).
Question 21 asks what month and year the breakup happened. We
are not interested in this information.
```{r}
HCMSTData <- HCMSTData[, !grepl("w1_q21", names(HCMSTData))]
```

Keep only one of each of the following features:

  * Education: keep "w1_ppeducat" and remove "w1_ppeduc"
  
  * Household income: keep "w1_ppincimp_cat" and remove "w2_log_real_inc" and "w1_ppincimp"
  
  * Partner age: keep "w1_ppagecat" and remove "w1_ppagect4" and "w1_ppage"
  
  * Partnership status: keep "w1_partnership_status" and remove "w1_section", "w1_partnership_status_cohab"
  and "xpartner_type_cohab"

  * Relationship quality: keep "w1_q34_reduced" and remove "w1_q34"

  * Race: keep w1_ppRace_* and remove w1_interracial_5cat
  
  * Sexual identity: keep "w1_identity_all" (more information) and remove "w1_identity_all_modified"
  
  * State of residence: keep "w1_PPREG4" and remove "w1_ppreg9"
  
```{r}
variablePairs <- c("w1_ppeduc", "w2_log_real_inc", "w1_ppincimp", "w1_ppagect4",
                   "w1_ppage", "w1_section", "w1_partnership_status_cohab",
                   "xpartner_type_cohab", "w1_q34", "w1_interracial_5cat",
                   "w1_identity_all_modified", "w1_ppreg9")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% variablePairs]
```

As the number of variables is large, we will use correlation analysis to find
highly correlated variables and potentially drop them.
```{r}
threshold <- 0.5
highly_correlated_pairs <- subset(highly_correlated,
                        abs(correlation_matrix[highly_correlated]) > threshold)

# Display highly correlated variable pairs
for (i in 1:nrow(highly_correlated_pairs)) {
  var1 <- rownames(correlation_matrix)[highly_correlated_pairs[i, 1]]
  var2 <- colnames(correlation_matrix)[highly_correlated_pairs[i, 2]]
  correlation <- correlation_matrix[highly_correlated_pairs[i, 1], 
                                    highly_correlated_pairs[i, 2]]
  cat(var1, "and", var2, "are highly correlated (Correlation:",
      correlation, ")\n")
}
```

The marital status (w1_ppmarit) is highly correlated with:
* w1_ever_marriedq21: whether the subject has ever been married
* w1_max_relation_status: highest status this relationship has achieved
Keep marital status and drop the other two.
```{r}
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% c("w1_ever_marriedq21", "w1_max_relation_status"))]
```

The race variables are one-hot encoded (1: white, 2: african-american, 3:american-indian, etc.)
so it is not surprising that they are negatively correlated. We will keep them as is.

subject_yrsed encodes the years of education of the subject, and ppeducat is a categorical
representation of the same variable. We will keep subject_yrsed and drop ppeducat.
```{r}
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% c("ppeducat", "q21d_year"))]
```

Finally, the household size is correlated to the presence of children.
We decide to create a new feature called w1_children_in_hh that is 1 if there are children
in the household and 0 otherwise. Also drop household size
```{r}
columnsChildren <- c("w1_PPT01", "w1_PPT25", "w1_PPT612", "w1_PPT1317")

HCMSTData$w1_children_in_hh <- 0
for (i in 1:nrow(HCMSTData)) {
  if (sum(HCMSTData[i, columnsChildren]) > 0) {
    HCMSTData[i, "w1_children_in_hh"] <- 1
  }
}
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% c("w1_PPT01", "w1_PPT25", "w1_PPT612", "w1_PPT1317", "w1_pphhsize"))]
```

# Missing Values

Remove features that have more than 90% of missing values
```{r}
missingValues <- sapply(HCMSTData, function(x) sum(is.na(x)))
missingValues <- missingValues[missingValues > 0]
missingValues <- missingValues[missingValues > 0.9 * nrow(HCMSTData)]
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% names(missingValues)]
length(missingValues)
missingValues
```

List features with more than 10% missing values
```{r}
missing_values <- sapply(HCMSTData, function(x) sum(is.na(x)))
missing_values <- missing_values[missing_values > 0.1 * nrow(HCMSTData)]
missing_values
```

Drop all except for w1_q5, which determines if it is a same-sex couple.
```{r}
missing_values <- missing_values[!(names(missing_values) %in% c("w1_q5"))]
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% names(missing_values))]
```

Replace all w1_q5 missing values with 0
```{r}
HCMSTData$w1_q5[is.na(HCMSTData$w1_q5)] <- 0
```

## Refuse to Answer

Values of -1 indicate that the subject refused to answer the question.
```{r}
minus_one_values <- sapply(HCMSTData, function(x) sum(x == -1))
minus_one_values <- minus_one_values[order(minus_one_values, decreasing = TRUE)]
minus_one_values
```

Feature w1_PERSNET_hom and w1_ppc21310 are related to internet access.
We will drop them as they are not relevant to the analysis.
```{r}
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% c("w1_PERSNET_hom", "w1_ppc21310"))]
```

For the remaining features, we will replace -1 with 0
```{r}
HCMSTData[HCMSTData == -1] <- 0
```

# Identify Feature Types

Rank the features by the number of distinct values
```{r}
distinct_values <- sapply(HCMSTData, function(x) length(unique(x)))
distinct_values <- distinct_values[order(distinct_values, decreasing = TRUE)]
distinct_values
```
The following features relate to the date the subject met their partner.
We will drop them as they are not relevant to the analysis.
"w1_year_fraction_met", "w1_met_month",
"w1_relate_duration_in2017_years", "w1_year_fraction_relstart",
"w1_relationship_start_calmonth", "w1_year_met_from_1990",
"w1_met_after_2009"

```{r}
dateMetCols <- c("w1_year_fraction_met", "w1_met_month",
                 "w1_relate_duration_in2017_years", "w1_year_fraction_relstart",
                 "w1_relationship_start_calmonth", "w1_year_met_from_1990",
                 "w1_met_after_2009")
HCMSTData <- HCMSTData[, !(names(HCMSTData) %in% dateMetCols)]
```

# Other

Drop all w1_q15 columns. They contain the state and country of the subject.
```{r}
HCMSTData <- HCMSTData[, !grepl("w1_q15", names(HCMSTData))]
```

Drop w1_days_access_internet_perwk, w1_number_people_met, w1_otherdate_dichotomous
```{r}
otherColumns <- c("w1_days_access_internet_perwk",
                  "w1_number_people_met",
                  "w1_otherdate_dichotomous")
HCMSTData <- HCMSTData[, !names(HCMSTData) %in% otherColumns]

```

# Save Data

Number of remaining features
```{r}
dim(HCMSTData)
```

Save the cleaned data that will be used for the regression analysis.
```{r}
write.csv(HCMSTData, file = "data/HCMST_clean.csv", row.names = FALSE)
```
