---
title: "STATS 412 Final Project EDA"
author: "Suyeon Betty Hwang, Mariano Aloiso"
date: "2023-12-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# 1. Importing Data and Necessary Packages

```{r}
#Importing Packages
library(dplyr)
library(ggplot2)
library(rlang)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(readr)
col <- brewer.pal(5, "Set2")
```

```{r}
#loading the data
df <-read.csv("HCMST_clean (1).csv")
howtheymet <- read.csv("howTheyMet.csv")
dim(df) # 3510   89 
dim(howtheymet)
```

Remove people who didn't answer w2_section (relationship status) question. In other words, remove people who are not in the 2020 survey.

```{r}
df<- df[!is.na(df$w2_section),]
dim(df)
```

Transforming the categorical variable (whether in a relationship or not) into a binary variable. 1: still in a relationship, 0: not in a relationship. "A" (still married), "B" (still partnered) are both considered still in a relationship.

```{r}
df$w2_section <- ifelse(df$w2_section %in% c(1, 3), 1, 0)
```

```{r}
#Selecting only a subset of data: only extract respondents who are in relationships:

#We will do this by making the Partner Status into Binary Category
df$w1_partnership_status_binary <- ifelse(df$w1_partnership_status %in% c(1,2), 1, 0)
df$w1_partnership_status_binary

table(df$w1_partnership_status_binary)

df$w1_partnership_status_binary[df$w1_partnership_status_binary==0] <- "Unpartnered"
df$w1_partnership_status_binary[df$w1_partnership_status_binary==1] <- "Partnered" 

barplot(prop.table(table(df$w1_partnership_status_binary)),col = col, main = "2017 Partnership Status ")

```

From now on, we will work only with a subset of data: only the respondents who are in relationships. In order to do this, we will extract and make a subset of data:

```{r}
subset_df <- df[df$w1_partnership_status %in% c(1,2), ]
single <- df[df$w1_partnership_status %in% c(3,4),]
dim(single) #648 x 89
dim(subset_df) #2862 x90
```

We will only work with 2862 observations who are in a relationship in Wave 1.

# 2. EDA

### Race and Gender Demography

```{r}
#Partner's Race Demography 

subset_df$w1_q6b[subset_df$w1_q6b==-1] <-"Refused"
subset_df$w1_q6b[subset_df$w1_q6b==1] <-"White"
subset_df$w1_q6b[subset_df$w1_q6b==2] <-"Black"
subset_df$w1_q6b[subset_df$w1_q6b==3] <-"Native American"
subset_df$w1_q6b[subset_df$w1_q6b==4] <-"Asian"
subset_df$w1_q6b[subset_df$w1_q6b==5] <-"Other"
barplot(prop.table(table(subset_df$w1_q6b)), col = col, main = "Race of Partners")
"Refused
 2,682 1 White
 314 2 Black or African American
 36 3 American Indian, Aleut, or
 Eskimo
 158 4 Asian or Pacific Islander
 189 5 Other (please specify) "


subset_df$w1_ppethm[subset_df$w1_ppethm==1] <-"White"
subset_df$w1_ppethm[subset_df$w1_ppethm==2] <-"Black"
subset_df$w1_ppethm[subset_df$w1_ppethm==3] <-"Other"
subset_df$w1_ppethm[subset_df$w1_ppethm==4]  <-"Hispanic"
subset_df$w1_ppethm[subset_df$w1_ppethm==5]  <-"Other"
barplot(prop.table(table(subset_df$w1_ppethm)), col = col, main = "Race of Respondents")
"2,503 1 White, Non-Hispanic
 311 2 Black, Non-Hispanic
 147 3 Other, Non-Hispanic
 438 4 Hispanic
 111 5 2+ Races, Non-Hispanic "

```

```{r}

subset_df$w1_ppgender[subset_df$w1_ppgender==1] <-"Male"
subset_df$w1_ppgender[subset_df$w1_ppgender==2] <-"Female"
table(subset_df$w1_ppgender) #oroginal: 976: 1131
barplot(table(subset_df$w1_ppgender), col = col, main = "Sex of the Respondents" )
```

```{r}
table(subset_df$w1_xlgb)
barplot(prop.table(table(subset_df$w1_xlgb)),col = col, main = "LGBTQ Sample Size" )
"400 1 LGB sample
 3,110 2 gen pop"
```

Almost same number of male and female respondents.

### Household Size

```{r}
table(subset_df$w1_pphhsize)
barplot(prop.table(table(subset_df$w1_pphhsize)), col = col, main = "Household Size for People Who Are In Relationships")

```

Most of the respondents seem to live in a household size of two, which means with a partner and without a child yet.

```{r}
table(df$w1_pphhsize)
barplot(prop.table(table(df$w1_pphhsize)), col = col, main = "Household Size for People Who Are In Relationships")
```

### Income Size

```{r}
#Income Size for Partnered People
prop.table(table(subset_df$w1_ppincimp_cat))
subset_df$w1_ppincimp_cat[subset_df$w1_ppincimp_cat == 1 ] <- "1. <40K"
subset_df$w1_ppincimp_cat[subset_df$w1_ppincimp_cat == 2 ] <- "2. $40K - $75K"
subset_df$w1_ppincimp_cat[subset_df$w1_ppincimp_cat == 3 ] <- "3.  $75K- $125K"
subset_df$w1_ppincimp_cat[subset_df$w1_ppincimp_cat == 4 ] <- "4. > $125K"
barplot(prop.table(table(subset_df$w1_ppincimp_cat)), col = col, main = "Income Size")


#Income Size for Single Population
single$w1_ppincimp_cat[single$w1_ppincimp_cat == 1 ] <- " 1. <40K"
single$w1_ppincimp_cat[single$w1_ppincimp_catt == 2 ] <- " 2. $40K - $75K"
single$w1_ppincimp_cat[single$w1_ppincimp_cat == 3 ] <- " 3. $75K- $125K"
single$w1_ppincimp_cat[single$w1_ppincimp_cat == 4 ] <- "4.  > $125K"
prop.table(table(single$w1_ppincimp_cat))
barplot(prop.table(table(single$w1_ppincimp_cat)), col = col, main = "SingleIncome Size")
```

Respondents' incomes come from evenly distributed brackets.

```{r}
income_category <- c(rep(""))
subset_df$w1_ppincimp_cat
incomesize <- data.frame(subset_df$w1_ppincimp_cat, single$w1_ppincimp_cat)
ggplot(incomesize, aes(fill=condition, y=value, x = ))
```

```{r}
prop.table(table(subset_df$w1_pprent))
prop.table(table(single$w1_pprent))
par(mfrow = c(1,2))
barplot(prop.table(table(subset_df$w1_pprent)), col = col, main = "Ownership of Residence")
"2,472 1 Owned or being bought by you or
 someone in your household
 932 2 Rented for cash
 106 3 Occupied without payment of
 cash rent"

barplot(prop.table(table(single$w1_pprent)), col = col, main = "Ownership of Residence for Single ")

```

Most of the subjects seem to own a property that they are living in, which could be an indicator of stability.

### Martial and Relationship Status

36% of the respondents are not married; 63% of the respondents are married.

```{r}
table(df$w1_ppmarit)
barplot(table(df$w1_ppmarit), col = col, main = "Martial Status of Respondents in 2017")
"1,986 1 Married
 138 2 Widowed
 328 3 Divorced
 69 4 Separated
 712 5 Never married
 277 6 Living with partner"

prop.table(table(df$w1_ppmarit))
```

Most respondents are either married.

```{r}
df$w1_max_relation_status[df$w1_max_relation_status==0] <-"Never Married or Cohab"
df$w1_max_relation_status[df$w1_max_relation_status==1] <-"Cohab But Not Married"
df$w1_max_relation_status[df$w1_max_relation_status==2] <-"Married"

barplot(prop.table(table(df$w1_max_relation_status)), col = col, main = "Highest Martial Status 2017")
```

```{r}
table(subset_df$w1_xlgb)
barplot(prop.table(table(subset_df$w1_xlgb)),col = col, main = "LGBTQ Sample Size" )
"400 1 LGB sample
 3,110 2 gen pop"

```

### Hometown

```{r}

table(subset_df$w1_PPREG4)
subset_df$w1_PPREG4[subset_df$w1_PPREG4==1] <- "Northeast"
subset_df$w1_PPREG4[subset_df$w1_PPREG4==2] <- "Midwest"
subset_df$w1_PPREG4[subset_df$w1_PPREG4==3] <- "South"
subset_df$w1_PPREG4[subset_df$w1_PPREG4==4] <- "West"

barplot(prop.table(table(subset_df$w1_PPREG4)), col = col, main = "Subject's Birth Region" )

"641 1 Northeast
 770 2 Midwest
 1,268 3 South
 831 4 West "

table(single$w1_PPREG4)
prop.table(table(subset_df$w1_PPREG4))
```

### 

```{r}
hist(subset_df$w1_q16, xlab = " Number of Relatives Seen" , col = col, main = "Number of Relatives Seen")
```

### Age

```{r}
summary(subset_df$w1_q9)
table(subset_df$w1_q9)
hist(subset_df$w1_q9, col = col, xlab = "age" , main = "Partner's Age" )

prop.table(table(subset_df$w1_ppagecat))
barplot(prop.table(table(subset_df$w1_ppagecat)), col = col, main = "Respondent's Age")

```

Partner's age seems to be normally distributed. Most of the partner's age was between 50 and 60.

```{r}
boxplot(subset_df$w1_age_when_met, col = col, main = "Age When They Met Their Partner" )
hist(subset_df$w1_age_when_met, col = col, main = "Age When They Met Their Partner" )

```

Most of the respondents have met their partner around their twenties.

```{r}
agegroup <- prop.table(table(subset_df$w1_ppagecat, subset_df$w2_section))
barplot(agegroup, col = col, legend.text = c("married", "Partnered, not married" , "Unpartnered ", "Never had a partner")) 
```

### Education Status

```{r}
#Respondent's education: 
#subset_df$w1_ppeducat <- factor(subset_df$w1_ppeducat, levels = c("Less Than High School", "High School", "Some College", "Bachelor's Degree or Higher"))
subset_df$w1_ppeducat[subset_df$w1_ppeducat==1] <- "Less Than High School"
subset_df$w1_ppeducat[subset_df$w1_ppeducat==2] <- "High School"
subset_df$w1_ppeducat[subset_df$w1_ppeducat==3] <- "Some College"
subset_df$w1_ppeducat[subset_df$w1_ppeducat==4] <- "Bachelor's Degree or Higher"
prop.table(table(subset_df$w1_ppeducat))
barplot(prop.table(table(subset_df$w1_ppeducat)), col = col, main = "Respondent's Education Level")

"286 1 Less than high school
 932 2 High school
 1,029 3 Some college
 1,263 4 Bachelor's degree or higher "
```

```{r}
#single_ed <- factor(single$w1_ppeducat, levels = c("Less Than High School", "High School", "Some College", "Bachelor's Degree or Higher")) #doesn't work. 

single$w1_ppeducat[single$w1_ppeducat==1] <- "Less Than High School"
single$w1_ppeducat[single$w1_ppeducat==2] <- "High School"
single$w1_ppeducat[single$w1_ppeducat==3] <- "Some College"
single$w1_ppeducat[single$w1_ppeducat==4] <- "Bachelor's Degree or Higher"

table(single$w1_ppeducat)
barplot(prop.table(table(single$w1_ppeducat)), col = col, main = "Single Education Level")

```

Most of the respondents have obtained Bachelor's degree or higher.

```{r}
par(mfrow = c(1,2))
boxplot(subset_df$w1_subject_mother_yrsed, col = col, main = "Partner's Mother's Educational Attainment" )
boxplot(subset_df$w1_partner_mother_yrsed, col = col, main = "Subject's Mother's Educational Attainment" )

```

It seems like the respondents' and partners' mother's education attainment is similar.

```{r}
prop.table(table(subset_df$w1_q12))
barplot(prop.table(table(subset_df$w1_q12)), col =col, main = "Partner Political Afflication" )

"38 -1 Refused
 510 1 Strong Republican
 165 2 Not Strong Republican
 412 3 Leans Republican
 1,025 4 Undecided/Independent/Other
 504 5 Leans Democrat
 210 6 Not Strong Democrat
 530 7 Strong Democrat"
```

The biggest number of politcal affliation of the partner is undecieded/independent next to strong republican.

```{r}
table(subset_df$w1_sex_frequency)
barplot(table(subset_df$w1_sex_frequency), col =col, main = "Sex Frequency ")

"170 -1 Refused
 86 1 Once a day or more
 368 2 3 to 6 times a week
 744 3 Once or twice a week
 519 4 2 to 3 times a month
 969 5 Once a month or less
 654 ."

boxplot(subset_df$w1_weekly_sex_frequency, col =col, main = "Weekly Sex Frequency ")

table(subset_df$w1w2_sex_freq_diff) #check this one
```

Most couples have sex once a week or once a month/less. Additionally, there wasn't any difference between sex frequency from 2017 and 2020.

### Income and Employment Status

```{r}
table(subset_df$w1_ppincimp_cat)
prop.table(table(subset_df$w1_ppincimp_cat))
barplot(prop.table(table(subset_df$w1_ppincimp_cat)), col = col, main = " Income Categories of Partner")
"1,001 1 <$40K
 848 2 $40K-$75K
 876 3 $75K-$125K
 785 4 >$125K "
```

Partners' incomes seem evenly distributed among four different income brackets.

```{r}

subset_df$w1_ppwork[subset_df$w1_ppwork==1] <- "Working as an Employee"
subset_df$w1_ppwork[subset_df$w1_ppwork==2] <- "Self-Employed"
subset_df$w1_ppwork[subset_df$w1_ppwork==3] <- "Not Working (Layoff)"
subset_df$w1_ppwork[subset_df$w1_ppwork==4] <- "Not Working- looking for work"
subset_df$w1_ppwork[subset_df$w1_ppwork==5] <- "Retired"
subset_df$w1_ppwork[subset_df$w1_ppwork==6] <- "Disabled"
subset_df$w1_ppwork[subset_df$w1_ppwork==7] <- "Other"

table(subset_df$w1_ppwork)
barplot(prop.table(table(subset_df$w1_ppwork)),col = col, main = "Partner Employment Status" )

"w1_ppwork_percent <- w1_ppwork %>% 
                    count(category) %>%
                    mutate(percentage = n/ sum(n)*100)"
 "
 1 Working - as a paid employee
 278 2 Working - self-employed
 11 3 Not working - on temporary
 layoff from a job
 178 4 Not working - looking for work
 663 5 Not working - retired
 184 6 Not working - disabled
 228 7 Not working - other "
```

Most of the respondent's partners are working as an employee and the second highest common employment status of the partner is retired.

```{r}
table(single$w1_ppwork)
barplot(prop.table(table(single$w1_ppwork)), col = col, main = "Single Employment Status")
```

### Relationship Status/ Quality

```{r}
#Relationship Status of the Entire Dataset: 
table(df$w1_partnership_status) 

barplot(prop.table(table(df$w1_partnership_status)), col = col, main = "2017 Relationship Status")
#1: married, 2: partnered, not married, 3: unpartnered, has had past partner, 4: never had a partner
```

```{r}
#Partner Status (Binary) 2020

table(df$w2_section) #all the 2017 data 

table(subset_df$w2_section)
subset_df$w2_section[subset_df$w2_section==0] <- "Not With the Same Partner"
subset_df$w2_section[subset_df$w2_section==1] <- "With the Same Partner" 

barplot(prop.table(table(df$w2_section)),col = col, main = "2020 Partnership Status for All Respondents from 2017")

barplot(prop.table(table(subset_df$w2_section)),col = col, main = "2020 Partnership Status for Respondents Who Were Partnered in 2017 ")
```

Most couples from 2017 are still with the same partner in 2020.

```{r}
subset_df$w1_q34[subset_df$w1_q34 ==1] <- "Excellent"
subset_df$w1_q34[subset_df$w1_q34 ==2] <- "Good"
subset_df$w1_q34[subset_df$w1_q34 ==3] <- "Fair"
subset_df$w1_q34[subset_df$w1_q34 ==4] <- "Poor"
subset_df$w1_q34[subset_df$w1_q34 ==5] <- "Very Poor"
table(subset_df$w1_q34)
barplot(prop.table(table(subset_df$w1_q34)),  col = col, main = "2017 Relationship Quality")
"1,706 1 Excellent
 884 2 Good
 202 3 Fair
 32 4 Poor
 23 5 Very Poor
 663 . "
```

Most of the respondents from 2017 in the study seem to be in an excellent and good relationships.

###How They Met Data Set

```{r}
dim(howtheymet) #2107 observations

#partnered subjects
howtheymet_1<- howtheymet[howtheymet$w1_partnership_status %in% c(1,2), ]

dim(howtheymet_1) #1716 observations

par(mfrow = c(2,3))
barplot(prop.table(table(howtheymet_1$friends)), col = col, main = "Met Through Friends")

barplot(prop.table(table(howtheymet_1$family)), col = col, main = "Met Through Family")

barplot(prop.table(table(howtheymet_1$online)), col = col, main = "Met Through Online")

barplot(prop.table(table(howtheymet_1$social_events)), col = col, main = "Met Through Social Events")

barplot(prop.table(table(howtheymet_1$work)), col = col, main = "Met Through Work")

barplot(prop.table(table(howtheymet_1$school)), col = col, main = "Met Through School")

```

```{r}

prop.table(table(howtheymet_1$online))
```

```{r}
#Couples who remained couples 
couples_who_stayed <- howtheymet_1 %>% filter(howtheymet_1$w2_section==1)
dim(couples_who_stayed) #1488

par(mfrow = c(2,3))
barplot(prop.table(table(couples_who_stayed$friends)), col = col, main = "Met Through Friends")

barplot(prop.table(table(couples_who_stayed$family)), col = col, main = "Met Through Family")

barplot(prop.table(table(couples_who_stayed$online)), col = col, main = "Met Through Online")

barplot(prop.table(table(couples_who_stayed$social_events)), col = col, main = "Met Through Social Events")

barplot(prop.table(table(couples_who_stayed$work)), col = col, main = "Met Through Work")

barplot(prop.table(table(couples_who_stayed$school)), col = col, main = "Met Through School")

```

```{r}

```

